Introducing statistics with \\texttt{intRo}

### 2014-11-25 student-paper-competition-2014/paper.Rmd
some rearranging and added thoughts

\begin{figure}[ht!]
\centering
\includegraphics[width=0.8\textwidth]{images/ui_annotate.png}
\caption{The five elements that comprise the \texttt{intRo} application: 1) top navigation, 2) side navigation, 3) options panel, 4) results panel, and 5) code panel.}
\label{fig:ui}
\end{figure}

@@

There are multiple modules within \texttt{intRo}, which perform specific data analysis tasks that employ an easy to use point-and-click interface. Each module is placed within a higher level category to further help students navigate the tool. The default modules included in \texttt{intRo} are found in table \ref{tab:modules}, and more modules can be easy added by an instructor, as detailed in section \ref{sec:extensible}.

\begin{table}
\centering
\begin{tabular}[ht!]{l|l|p{.65\textwidth}}
Category & Module & Description \\
\hline
\multirow{2}{*}{Data} & Sources & The data sources module allows the user to select or upload a dataset. Several datasets are pre-included, which can be selected via a drop-down menu. The user can check a box to upload a dataset, and then select the CSV file from his computer. Upon selection or successful upload of a dataset, a table view of the data will be visible on the right. From this module the user can select a subset of the data either by value or random selection. The dataset selected in the data module will be made available to all other modules of intRo.\\

& Transformation & The data transformation modules allows transformation of numeric variables to categorical and categorical variable to numeric. Additionally, numeric variable can be transformed using a power transformation. The transformed variables can be saved to the dataset for use elsewhere in \texttt{intRo}.\\
\hline
\multirow{2}{*}{Summaries} & Graphical & The graphical summaries module provides several common graphical displays of variables in the dataset. First, the user selects the plot desired, and either a single variable or two variables from the dataset, depending on plot type. The variable available to choose will only be those compatible with plot type selected. The plot visible on the right will automatically update when these values are selected.\\

& Numerical & The numeric summaries module is another method of computing summaries of the data. The user is presented with a list of each variable from their dataset. The user can select one or more different variables, and will be presented with a table of some summary statistics. The summary statistics displayed vary depending on the type of the variables. For instance, for a numeric variable, a five number summary will be displayed. Character or factor variables will display counts of each unique value of that particular variable. Additionally, the user has the ability to display summaries by a grouping variable.\\
\hline
\multirow{3}{*}{Statistics} & Contingency & The contingency table module allows the user to compute a contingency table for two categorical variables. The results can either be displayed as counts, row percentages, column percentages, or total percentages.\\

& Regression & The simple linear regression module allows the user to compute a simple linear regression of a dependent variable (y) on an independent variable (x). The module will automatically filter any variables that are non-numeric so that the user may only select variables that are suitable for a simple linear regression. Results of the regression will be displayed on the right. A table of parameter estimates, a scatterplot of the data along with a line of best fit, The R-Squared value, and residual diagnostic plots will all be displayed. A user also has the ability to save residuals and predicted values to their dataset for further analysis.\\

& T test & The T test module allows the user to perform either a one variable or two variable Student's T test. The user can select configuration options, including the direction of the test and the confidence level of the test. The results, including the t statistic, degrees of freedom, p-value, and a confidence interval, will be displayed in a text box in the results panel.
\end{tabular}
\caption{A list of default modules that ship with \texttt{intRo} as well as brief descriptions of each.}
\label{tab:modules}
\end{table}

## Exciting
In general, introduction to statistics is a class undergraduates dread. It's hard, the language is weird, and the software is hard to use. With \texttt{intRo}, we hope to flip this paradigm. \texttt{intRo} is a fun, easy to use tool that's available on the web, and is even usable on mobile devices. As a web-based application, this tool is immediately more familiar to students than a desktop application. With interactive plots from \texttt{ggvis} [@ggvis], \texttt{intRo} brings a `wow'-factor to learning introductory statistics.

Once students get pulled in by an easy to use interface and fun interaction with their data, \texttt{intRo} has an ulterior motive: to get students excited about programming. By navigating about the user interface of \texttt{intRo}, students are actually creating a fully-executable \texttt{R} script that they can download and run locally. This code creation element of \texttt{intRo} is meant to generate excitement about programming in \texttt{R} and empower students to feel that they can generate code as well.

@@

@@

Reactive programming is a programming paradigm that ``tackles issues posed by event-driven applications by providing abstractions to express programs as reactions to external events and having the language automatically manage the flow of time (by conceptually supporting simultaneity), and data and computation dependencies.'' [@reactive2013] As implemented by \texttt{shiny} [@shiny], results automatically update when users interact with the interface.

@@

In the creation of \texttt{intRo} we walked a fine line between giving the user flexibility and having realistic usability. At the same time, \texttt{intRo} was created as a consumer of another package, \texttt{shiny}, in which we as developers were the beneficiaries of another team of developers' decision to balance flexibility and usability. For a tangible example, consider the graphical summaries module. We only allow variables of a type consistent with the selected plot to be displayed. This is a conscious decision that limits an \texttt{intRo} user's flexibility, while maximizing the usability (by minimizing crashes) of the application. On the flip side of this, \texttt{shiny} allows much higher flexibility. For instance, the entire application (including user interface) is created dynamically upon load, based on the modules currently housed within \texttt{intRo}. [Maybe add a figure here?] However, \texttt{shiny} does have limits in its flexibility based on the designers decisions for usability. One current example the slider element. This element allows for fixed width steps from its minimum to its maximum. The JavaScript library being utilized in \texttt{shiny} allows for arbitrary function calls to to generate these steps, however they much be written in plain JavaScript. This is an example of a decision made by the developers of \texttt{shiny} to limit functionality in favor of usability of their package.

@@

>>>>

\begin{figure}[ht!]
\centering
\includegraphics[width=0.8\textwidth]{images/ui_annotate.png}
\caption{The five elements that comprise the \texttt{intRo} application: 1) top navigation, 2) side navigation, 3) options panel, 4) results panel, and 5) code panel. \ak{[Don't forget to update this.]}}
\label{fig:ui}
\end{figure}

@@

There are multiple modules within \texttt{intRo}, which perform specific data analysis tasks that employ an easy to use point-and-click interface. Each module is placed within a higher level category to further help students navigate the tool. The default modules included in \texttt{intRo} are split into three higher level categories - data, summaries, and statistics - and more modules can be easy added by an instructor, as detailed in section \ref{sec:extensible}.

Within the data category there are two modules, sources and transformation. The data sources module allows the user to select or upload a dataset. Several datasets are pre-included, which can be selected via a drop-down menu. The user can check a box to upload a dataset, and then select the CSV file from his computer. Upon selection or successful upload of a dataset, a table view of the data will be visible on the right. From this module the user can select a subset of the data either by value or random selection. The dataset selected in the data module will be made available to all other modules of intRo. The data transformation module allows transformation of numeric variables to categorical and categorical variable to numeric. Additionally, numeric variable can be transformed using a power transformation. The transformed variables can be saved to the dataset for use elsewhere in \texttt{intRo}.

The graphical summaries module provides several common graphical displays of variables in the dataset. First, the user selects the plot desired, and either a single variable or two variables from the dataset, depending on plot type. The variable available to choose will only be those compatible with plot type selected. The plot visible on the right will automatically update when these values are selected. The numeric summaries module is another method of computing summaries of the data. The user is presented with a list of each variable from their dataset. The user can select one or more different variables, and will be presented with a table of some summary statistics. The summary statistics displayed vary depending on the type of the variables. For instance, for a numeric variable, a five number summary will be displayed. Character or factor variables will display counts of each unique value of that particular variable. Additionally, the user has the ability to display summaries by a grouping variable.

Within the set statistics modules, there are three simple statistics tasks that an introductory statistics student will learn in his first semester. The contingency table module allows the user to compute a contingency table for two categorical variables. The results can either be displayed as counts, row percentages, column percentages, or total percentages. The simple linear regression module allows the user to compute a simple linear regression of a dependent variable (y) on an independent variable (x). The module will automatically filter any variables that are non-numeric so that the user may only select variables that are suitable for a simple linear regression. Results of the regression will be displayed on the right. A table of parameter estimates, a scatterplot of the data along with a line of best fit, The R-Squared value, and residual diagnostic plots will all be displayed. A user also has the ability to save residuals and predicted values to their dataset for further analysis. The T test module allows the user to perform either a one variable or two variable Student's T test. The user can select configuration options, including the direction of the test and the confidence level of the test. The results, including the t statistic, degrees of freedom, p-value, and a confidence interval, will be displayed in a text box in the results panel.

\ak{[Add blurb about easy to read/consistent documentation]}

## Exciting
In general, introduction to statistics is a class undergraduate students dread. It's hard, the language is unfamiliar, and the software is hard to use. With \texttt{intRo}, we hope to flip this paradigm. \texttt{intRo} is a fun, easy to use tool that's available on the web, and is also usable on mobile devices. As a web-based application, this tool is immediately more familiar to students than a desktop application. With interactive plots from \texttt{ggvis} [@ggvis], \texttt{intRo} brings a `wow'-factor to learning introductory statistics.

Once students get pulled in by an easy to use interface and fun interaction with their data, \texttt{intRo} has an ulterior motive: to get students excited about programming. By navigating about the user interface of \texttt{intRo}, students are actually creating a fully-executable \texttt{R} script that they can download and run locally as well as viewing the script change real-time within the application. This code creation element of \texttt{intRo} is meant to generate excitement about programming in \texttt{R} and empower students to feel that they can generate code as well.

@@

For example, an instructor may wish to introduce hypothesis testing from a randomization framework rather than distributional. In this case, a non-parametric test could be added easily as a module to be used instead of the T test. This addition would be fairly simple to complete. The user would fork \texttt{intRo} from http://github.com/gammarama/intRo and begin by adding six types of files to a folder within a module folder: `helper.R`, `observe.R`, `output.R`, `reactive.R`, `static.R`, and `ui.R`. Each of these files is written to comply with the reactive nature available in a \texttt{shiny} application. The `ui.R` file contains all the user interface elements within the new module's tab. This will include any options in the options panel as well as results. The `helper.R` and `static.R` files refer to static functions and global variable that may be used within the module. The file `output.R` contains the reactive elements that tie output to the user interface within \texttt{shiny} and `observe.R` and `reactive.R` contain elements within an `observer` function or `reactive` function within the \texttt{shiny} framework. By placing each of these files within the `modules/nonparametric_test` folder and adding a single line to a navigation file indicating under which module grouping the non-parametric test module should go, a new module is available within \texttt{intRo} upon running the application.

@@

Reactive programming is a programming paradigm that ``tackles issues posed by event-driven applications by providing abstractions to express programs as reactions to external events and having the language automatically manage the flow of time (by conceptually supporting simultaneity), and data and computation dependencies.'' [@bainomugisha2012survey] As implemented by \texttt{shiny} [@shiny], results automatically update when users interact with the interface.

@@

\ak{[Add information about save buttons for individual pieces of code]}

In the creation of \texttt{intRo} we walked a fine line between giving the user flexibility and having realistic usability. At the same time, \texttt{intRo} was created as a consumer of another package, \texttt{shiny}, in which we as developers were the beneficiaries of another team of developers' decision to balance flexibility and usability. For a tangible example, consider the graphical summaries module. We only allow variables of a type consistent with the selected plot to be displayed. This is a conscious decision that limits an \texttt{intRo} user's flexibility, while maximizing the usability (by minimizing crashes) of the application. On the flip side of this, \texttt{shiny} allows much higher flexibility. For instance, the entire application (including user interface) is created dynamically upon load, based on the modules currently housed within \texttt{intRo}. \ak{[Maybe add a figure here?, including example code for dynamically loading the ui]} However, \texttt{shiny} does have limits in its flexibility based on the designers decisions for usability. One current example the slider element. This element allows for fixed width steps from its minimum to its maximum. The JavaScript library being utilized in \texttt{shiny} allows for arbitrary function calls to to generate these steps, however they mustf be written in plain JavaScript. This is an example of a decision made by the developers of \texttt{shiny} to limit functionality in favor of usability of their package.

@@

\ak{[Make sure we talk about the upcoming R package for module creation, mention also having intro in the classroom in the spring]}
\ak{[Future challenges: server performance?]}
